---
title: "A Glimpse of Your Data"
author: "Yun Dai"
date: "01/2019"
output: 
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    toc_float: true
    css: contents.css
---

******
## Summary Statistics
`head()`, `str()`, `summary()` and `table()` are useful functions to get a sense of our data in terms of the structure and summary statistics of the objects.

******
### head(), tail()
`head()` returns the first several rows of an object. The default number of rows is 6.
```{r cars}
head(mtcars)
```

We can set how many rows to be displayed. 
```{r}
head(mtcars, n = 10)
```

The `tail()` function, alternatively, returns the last several rows of the object.
```{r}
tail(mtcars, 2)
```


******
### str()
The `str()` function displays the structure of an object.
```{r iris}
str(iris)
```

We can also go to the level of vector.
```{r}
str(iris$Species)
```

******
### summary()
`summary()` provides us with the basic summary statistics.

Summarizing the dataset:
```{r}
summary(iris)
```

Summarizing a column:
```{r}
summary(iris$Species)
```

`summary()` will also tell us the number of missing values, if there are any.
```{r}
library(dplyr)
summary(storms$ts_diameter)
```


******
### table()
`table()` gives us a frequency table.
```{r}
table(iris$Species)
```

`table()` can be more useful in cross-tabulation.
```{r}
table(iris$Species, iris$Petal.Width)
```


******
### listing cases
To examine a subset of data, we can print the cases in the R console by [subsetting the data frame](https://shanghai.hosting.nyu.edu/data/r/subsetting.html#subsetting_data_frames).

Rows (cases) and columns (variables):
```{r}
mtcars[10:15, 2:5]
```

Conditional:
```{r}
mtcars[mtcars$mpg > 25, ]
```

******
## Exploratory Analysis
`cor()`, `t.test()` and `lm()` are useful functions for exploratory analysis.

******
### correlations
`cor()` produces correlations.
```{r}
cor(mtcars[,1:3])
```

`use` argument allows us to decide if we want to apply pairwise deletion of missing values should we have missing data, for instance. 
```{r}
a <- mtcars
a[2, 1] <- NA
a[5, 3] <- NA

cor(a[,1:3], use = "pairwise.complete.obs")
```

Or listwise deletion of missing values.
```{r}
cor(a[,1:3], use = "complete.obs")
```


******
### t-tests
`t.test()` performs one and two sample t-tests.

Let's create a group variable by some arbitrary standard first with the `mtcars` data before performing a t-test.
```{r}
mtcars$group <- NA
mtcars$group[mtcars$cyl > 4] <- 1
mtcars$group[mtcars$cyl <= 4] <- 2
```

Now we can perform an independent 2-group t-test by `group`. 
```{r}
t.test(mpg ~ group, data = mtcars)
```

We can also do paired t-test and one sample t-test with `t.test()`.

******
### linear regressions
`lm()` fits linear models. Below are some arbitrary examples of using `lm()` to fit linear regressions. `+` indicates multiple independent variables. `:` indicates interactions without main effects, and `*` indicates interactions with main effects.
```{r}
model1 <- lm(mpg ~ cyl, data = mtcars)
model2 <- lm(mpg ~ cyl + disp + vs, data = mtcars)
model3 <- lm(mpg ~ cyl : disp, data = mtcars)
model4 <- lm(mpg ~ cyl * disp, data = mtcars)
```

`summary()` returns the summary statistics of the model.
```{r}
summary(model2)
```

We can also obtain the model components of interest to us, which are stored in a list.
```{r}
summary(model2)$coefficient
```

```{r}
summary(model2)$residuals
```

