---
title: "A Lot of Bars"
author: "Yun Dai"
date: "08/2018"
output: 
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    toc_float: true
    css: contents.css
---

******
This post was adapted from a section in [The Jigsaw Puzzle Pieces - Creating the Graphs with ggplot2](https://shanghai.hosting.nyu.edu/data/r/case-1-2-ggplot2.html), Part II of a series of posts on building Shiny App. We created many bar plots there and the variations of bar plots. For summary purposes, I want to bring all of them together here.

## First Steps
As usual, before we begin, let's load all the packages we will need.
```{r warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
```

### Sample data 
First we'll load the sample survey data.
```{r message=FALSE, warning=FALSE}
## setwd()
load("/Volumes/GoogleDrive/My Drive/websites/new web/data/r/sample/survey")

head(survey)
```

### Color palette 
We'll create a list of a color palette. This will make working with colors more easily later in plotting.
```{r message=FALSE, warning=FALSE}
palette <- list(purple = c("#351F39", "#351C4D", "#6c1f55", "#765285", "#8a6899"),
                turquoise = c("#709FB0", "#849974", "#A0C1B8"),
                golden = c("#D1A827", "#f3da4c"))
```

******
## A Lot of Bars
Let's say we want to present how often respondents of different majors use the library website to search for articles. The data on website usage is `Q1.3` ("I use our library website to (how often) - Search for articles"). The data for major is `major`. 

`Q1.3` and `major` both contain discrete values. Bar plots and their variations can easily achieve this goal. 
```{r message=FALSE, warning=FALSE}
tail(survey[c("Q1.3","major")])
```

```{r message=FALSE, warning=FALSE}
table(survey$Q1.3)
```

```{r message=FALSE, warning=FALSE}
table(survey$major)
```

******
### Stacked bars (count)
```{r message=FALSE, warning=FALSE}
ggplot(survey, aes(major)) + 
  geom_bar(aes(fill = Q1.3), position = position_stack(reverse = TRUE), width = 0.4, alpha = 0.75) +
## geom_bar() adds a layer of stacked bars to the plot
## aes(fill = Q1.3) fills the stacked bars with the counts of each value of Q1.3 (never/occasionally/often)
## position = position_stack(reverse = TRUE) reverses the order of the stacked bars
  scale_fill_manual(values = c(palette[[1]][4], palette[[2]][1], palette[[3]][1])) +
## manually fills the bars with the preset color scheme
  scale_x_discrete(limits = rev(levels(survey$major))) +
## reverse the order of levels of x axis  
  coord_flip() +
## flip the coordinates
theme(axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12, margin = margin(0,3,0,0)),
      axis.title.y = element_blank(),
      axis.title.x = element_blank(),
      axis.ticks.x = element_line(size = 0),
      legend.title = element_blank(),
      legend.text = element_text(size = 10),
      plot.margin = unit(c(1,1,1,1), "cm")) 
```

******
### Stacked bars (percentage)
Alternatively, we can fill the stacked bars with the percentage of each value of `Q1.3`.
```{r message=FALSE, warning=FALSE}
ggplot(survey, aes(major)) + 
  geom_bar(aes(fill = Q1.3), position = "fill", width = 0.4, alpha = 0.75) +
## position = "fill" sets the plot as stacked bars with filled percentages instead of counts
  scale_fill_manual(values = c(palette[[1]][4], palette[[2]][1], palette[[3]][1])) +
  scale_x_discrete(limits = rev(levels(survey$major))) +
  coord_flip() +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12, margin = margin(0,3,0,0)),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_line(size = 0),
        legend.title = element_blank(),
        legend.text = element_text(size = 10),
        plot.margin = unit(c(1,1,1,1), "cm")) 
```

******
### Grouped bars
Or we can group the bars rather than stack the bars.

```{r message=FALSE, warning=FALSE}
ggplot(survey, aes(major)) + 
  geom_bar(aes(fill = Q1.3), position = "dodge", width = 0.4, alpha = 0.75) +
## position = "dodge" sets the plot as grouped bars
  scale_fill_manual(values = c(palette[[1]][4], palette[[2]][1], palette[[3]][1])) +
  scale_x_discrete(limits = rev(levels(survey$major))) +
  coord_flip() +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12, margin = margin(0,3,0,0)),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_line(size = 0),
        legend.title = element_blank(),
        legend.text = element_text(size = 10),
        plot.margin = unit(c(1,1,1,1), "cm")) 
```

******
### Lollipop chart
Bar charts are cognitively light to process. But we could also be easily bored by bar charts. A lollipop chart, which is a variation of bar chart, can do the same thing for us with some visual diversity.

To make a lollipop chart, we need a frequency table that summarizes the counts of each major. 
```{r message=FALSE, warning=FALSE}
tb <- survey %>% count(major) %>% data.frame() # generates a frequency table
tb <- tb %>% arrange(-n) %>% filter(major != "Undefined") # reorder by frequency; remove undefined majors
tb
```

Now let's plot the lollipop chart from the frequency table. 
```{r message=FALSE, warning=FALSE}
ggplot(tb, aes(n, reorder(major, -n), label = n)) +
  geom_segment(aes(x = 0, y = reorder(major, -n), xend = n, yend = reorder(major, -n)), 
               size = 0.5, color = "grey50") +
  ## reorder Group2 by its frequency in descending order
  geom_point(size = 8) +
  ## geom_point() creates scatterplots that display the relationship between two variables, which takes the values of both x and y axes.
  ## Here, if we add geom_point() before geom_segment(), we'll see the sticks stabbing the points. 
  ## So the sequence actually matters here. 
  geom_text(color = "white", size = 3) +
  coord_flip() +
  theme(axis.text.x = element_text(size = 12, angle = 90, hjust = 1),
  ## The label of x axis would be too wide to lay out horizontally. hjust = 1 means right-justified.
        axis.text.y = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_line(size = 0))
```



