---
title: "Aggregating Data to Get Summaries"
author: "Yun Dai"
date: "07/2019"
output: 
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    toc_float: true
    css: contents.css
---


```{r echo=FALSE, message=FALSE, warning=FALSE}
rm(list = ls())
load("/Volumes/GoogleDrive/My Drive/websites/new web/data/r/sample/session2")
library(dplyr)
session <- session2 %>% select(-event) %>% ungroup()
```

******
In this post, we demonstrate two ways of aggregating and summarizing data: with package `dplyr` and with package `data.table`.

******
First, let's load the sample dataset `session`, which has recorded how much time a user has spent on each resource type on which device type on an online course platform.

```{r}
head(session)
```

```{r}
summary(session$resource)
```

```{r}
summary(session$device)
```

```{r}
summary(session$interval)
```


******
## `dplyr` functions

There are several functions in `dplyr` that aggregate data for summaries: `tally()`, `count()`, and `summarise()`. Below we explain how each one works with examples.

```{r message=FALSE,warning=FALSE}
library(dplyr)
```

******
### `tally()`

`tally()` calls `n()` to count the number of cases or calls `sum()` to calculate the total amount. 

******
#### `tally(n())`
Below we count the number of cases by the resource type. Note that `n()` does not take arguments.

```{r}
session %>% filter(!is.na(resource)) %>% group_by(resource) %>% tally(n())
```

******
#### `tally(sum())`

`tally(sum())` calculates the sum by the specified group. 

Below we sum the time spent on each resource type in minutes. 
```{r}
session %>% filter(!is.na(resource)) %>% group_by(resource) %>% tally(sum(interval) / 60)
```

******
#### `add_tally()`

`add_tally()` is short-hand for `mutate()` and adds a column named `n` to the data frame.

Below column `n` is the sum of the time spent on each `resource` type by `date` and `resource`.

```{r}
session %>% filter(!is.na(resource)) %>% group_by(date, resource) %>% add_tally(sum(interval) / 60)
```


******
### `count()`

`count()` is similar to `tally()`, but calls `group_by()` before and `ungroup()` after. If the data is already grouped, `count()` adds an additional group that is removed afterwards.

Here, using `count()` to get the number of cases by each resource type returns the same result as with `tally()` above. With `count(resource)`, data is grouped by `resource`.
```{r}
session %>% filter(!is.na(resource)) %>% count(resource)
```

******
In the example below, we count the number of cases grouped by both `id` and `resource`.
```{r}
session %>% filter(!is.na(resource)) %>% count(id, resource)
```




******
#### `add_count()`

`add_count()` is a short-hand for `group_by()` + `add_tally()`. 

Below column `n` is the count of `resource` by `date` and `resource`.
```{r}
session %>% filter(!is.na(resource)) %>% add_count(date, resource)
```


******
### `summarise()`
`summarise()` reduces multiple values down to a single value. The summary functions include measures for center (`mean()`, `median()`), spread (`sd()`, `IQR()`, `mad()`), range (`min()`, `max()`, `quantile()`), position (`first()`, `last()`, `nth()`), count (`n()`, `n_distinct()`) and logical evaluations (`any()`, `all()`) of data. 

Together with `group_by()`, `summarise()` returns summaries by the specified group. Without grouping, `summarise()` returns a single row.

Below we calculate the mean of the total time (in minutes) spent on each `resource` by `date` and `resource`.
```{r}
session %>% 
  filter(!is.na(resource)) %>% 
  group_by(date, resource) %>% 
  tally(sum(interval)) %>% 
  group_by(resource) %>%
  summarise(mean(n) / 60)
```

******
Without grouping data, below we get a single row, the sum of the time spent on all resources.
```{r}
session %>% 
  filter(!is.na(resource)) %>% 
  tally(sum(interval / 60)) 
```

******
## `data.table` functions

`data.table` offers efficient methods to aggregate and summarize data with the `data.table` syntax `DT[i, j, by]`. `i` indicates on which rows; `j` indicates what to do; `by` indicates being grouped by what.

```{r message=FALSE,warning=FALSE}
library(data.table)

session <- data.table(session)
```

******
### count
Below we first count the number of cases by the resource type. 

`.N` is a `data.table` [special-symbol](https://www.rdocumentation.org/packages/data.table/versions/1.10.0/topics/special-symbols). It is an integer containing the number of rows in the group, and is renamed to `N` in the result. `.N` may be useful when the column names are not known in advance. 

```{r}
session[!is.na(resource), .N, by = resource]
```

It is equivalent to the `dplyr` way of 
```{r eval=FALSE}
session %>% filter(!is.na(resource)) %>% group_by(resource) %>% tally(n())
```

******
### sum

Below we sum the time spent on each resource type in minutes. 

```{r}
session[!is.na(resource), sum(interval)/60, by = resource]
```

It is equivalent to the `dplyr` way of 
```{r eval=FALSE}
session %>% filter(!is.na(resource)) %>% group_by(resource) %>% tally(sum(interval) / 60)
```


******
### adding a column 

Like with `dplyr::add_tally()`, we can create a new column with `:=` in the data table that contains the summaries instead of aggregating the data table. 

Below column `sum` is the sum of the time spent on each `resource` type by `date` and `resource`. 

Note that the `data.table` method keeps the missing cases of `resource` in the result, while the `dplyr` way excludes the rows where `resource` is `NA`.
```{r}
session[!is.na(resource), sum := sum(interval)/60, by = c("date", "resource")]
session
```

It is equivalent to the `dplyr` way of 
```{r eval=FALSE}
session %>% filter(!is.na(resource)) %>% group_by(date, resource) %>% add_tally(sum(interval) / 60)
```


******
Below we calculate the mean of the total time (in minutes) spent on each `resource` by `date` and `resource`.

```{r}
s <- session[!is.na(resource), sum(interval), by = c("date", "resource")]
s[, mean(V1)/60, by = resource]
```

It is equivalent to the `dplyr` way of 

```{r eval=FALSE}
session %>% 
  filter(!is.na(resource)) %>% 
  group_by(date, resource) %>% 
  tally(sum(interval)) %>% 
  group_by(resource) %>% 
  summarise(mean(n)/60)
```

