---
title: "Part II: The Shiny App"
author: "Yun Dai"
date: "04/2019"
output: 
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    toc_float: true
    css: contents.css
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
rm(list = ls())
load("/Volumes/GoogleDrive/My Drive/websites/new web/data/r/sample/demo4/map")
load("/Volumes/GoogleDrive/My Drive/websites/new web/data/r/sample/demo5/resource_day_sum")
load("/Volumes/GoogleDrive/My Drive/websites/new web/data/r/sample/demo6/local")

names(local)[2] <- "localhour"
```


******
This is the second post following the [the previous one](https://shanghai.hosting.nyu.edu/data/r/case-2-1-animation.html), where we used tracking logs from an online course platform to create animated visualizations. 

In this post, we explore three aspects of the user engagement: (1) global distribution of the users over time, (2) resource usage over time, and (3) the peak hours of accessing the platform by device types (desktop versus mobile).

With the capabilities of the [slider](http://shiny.rstudio.com/gallery/slider-bar-and-slider-range.html), we are able to pick a certain date or select a date range to explore the graphs interactively; we can also [run animation sequences](https://shiny.rstudio.com/articles/sliders.html) with the slider controls.

******
## Global distributions 
We use the same data as in the [earlier post](https://shanghai.hosting.nyu.edu/data/r/case-2-1-animation.html#global_distributions), where we have created an animated graph with `gganimate`.

```{r}
head(map)
```

Here we want to (1) view the map on a selected date, and (2) run the animation sequences once we hit the play button, as shown below.

<iframe src ="https://yundai.shinyapps.io/demo4/" height= 600px width = 900px />

To create something as shown above, on the `ui` side, for the slider, we need to specify the display range for the `sliderInput()`: `min` sets the minimum value to use in the slider bar, and `max` sets the maximum value; both should be date objects in our case. With the `animate` argument and the `interval` option, we set the pace of the iterations in the animation. We set the `step` to be 1 so that the increment between the values will be one day and in the animation the map will reflect the change daily. `value` sets the default value to show in the slider bar when the app is first loaded.

```{r eval=FALSE}
library(shiny)
library(ggthemes)
library(maps)
library(ggplot2)
library(dplyr)

load("map")

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      sliderInput(inputId = "distribution", 
                  label = "Dates:",
                  min = as.Date("2016-01-24","%Y-%m-%d"),
                  max = as.Date("2016-04-02","%Y-%m-%d"),
                  value = as.Date("2016-02-29"), timeFormat="%Y-%m-%d", 
                  step = 1,
                  animate = animationOptions(interval = 1800))),
    mainPanel(plotOutput(outputId = "map", height = "70vh"))))
```

On the `server` side, we plot the graph on the subset of data whose `date` range is within the `min` and `max` values by setting `date == input$distribution`. `distribution` is the `inputId` of the slider.

```{r eval=FALSE}
server <- function(input, output){
  output$map <- renderPlot({
    map %>% 
      filter(date == input$distribution) %>%
      ggplot() +
      borders("world", colour = "gray90", fill = "gray85") +
      theme_map() + 
      geom_point(aes(x = longitude, y = latitude, size = n), 
                 colour = "#351C4D", alpha = 0.55) +
      labs(size = "Users") + 
      ggtitle("Distribution of Users Online") 
  })
}

shinyApp(ui = ui, server = server)
```

******
## Resource usage trajectory
Not only can we input single values in the slider, but also ranges. To explore how users accessed the platform within a date range, we customize the slider so that we can extract data within a period of time. 

<iframe src ="https://yundai.shinyapps.io/demo5/" height= 600px width = 900px />

Below is the sample data, same as what we used to create the [animated line graph](https://shanghai.hosting.nyu.edu/data/r/case-2-1-animation.html#resource_usage_trajectory).
```{r}
head(resource_day_sum)
```

On the `ui` side, same as what we have done in the section above, we need to set the `sliderInput()` for display range with arguments `min` and `max`.
```{r eval=FALSE}
library(shiny)
library(ggplot2)
library(dplyr)

load("resource_day_sum")

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      sliderInput(inputId = "trajectory", 
                  label = "Date Range:",
                  min = as.POSIXct("2016-01-24","%Y-%m-%d"),
                  max = as.POSIXct("2016-04-02","%Y-%m-%d"),
                  value = c(as.POSIXct("2016-02-01"), as.POSIXct("2016-03-21")), 
                  timeFormat="%Y-%m-%d", step = 1)),
    mainPanel(plotOutput(outputId = "lines", height = "70vh"))))
```

On the `server` side, to match the date range we set for the slider, we plot the graph on the subset of data where `date >= input$trajectory[1] & date <= input$trajectory[2]`. `trajectory` is the `inputId` of the slider.
```{r eval=FALSE}
server <- function(input, output){
  color <- c("#765285","#D1A827","#709FB0", "#849974", "#A0C1B8")
  due <- c(as.POSIXct("2016-01-25 UTC"), as.POSIXct("2016-02-01 UTC"), 
         as.POSIXct("2016-02-15 UTC"), as.POSIXct("2016-02-22 UTC"), 
         as.POSIXct("2016-02-29 UTC"), as.POSIXct("2016-03-07 UTC"),
         as.POSIXct("2016-03-14 UTC"), as.POSIXct("2016-03-21 UTC"), 
         as.POSIXct("2016-03-28 UTC"))

  output$lines <- renderPlot({
    resource_day_sum %>%
      filter(date >= input$trajectory[1] & date <= input$trajectory[2]) %>% 
      ggplot(aes(date, sum, group = resource, colour = resource)) + 
      geom_line(alpha = 0.9, size = 0.65) + 
      scale_x_datetime(breaks = seq(as.POSIXct("2016-01-26 UTC"), as.POSIXct("2016-04-02 UTC"), "7 days"), date_labels = "%b %d") +
      geom_vline(xintercept = due, alpha = 0.6, size = 0.65, colour = "grey55") +
      scale_colour_manual(values = color, name = "Resource") +
      labs(title = "Time Spent Online Learning", x = "Date", y = "Total Minutes per Day")
  })
}

shinyApp(ui = ui, server = server)
```


******
## Peak hours by device access
Finally, we would like to create a heatmap to reveal the peak hours of users accessing the online resources. We showed how to make a [static heatmap](https://shanghai.hosting.nyu.edu/data/r/local-time.html#plotting_peak_hours) previously. Here we add the feature that enables people to interactively view the heatmap by device types. 

<iframe src ="https://yundai.shinyapps.io/demo6/" height= 600px width = 900px />

We have added the device data to the data of local hours.
```{r}
head(local, 15)
```

The tick marks on the X axis indicate the due dates. Clearly, when the due dates approached, there were more students accessing online resources, ususally lasting from evening to midnight. Inspecting the peak hours by device types, it is apparent that the majority accessed the online resources via desktop devices, while very few accessed the resources via mobile devices. 

We have shown how to build a simple dashboard like this [before](https://shanghai.hosting.nyu.edu/data/r/case-1-3-shinyapp.html).
```{r eval=FALSE}
library(shiny)
library(shinythemes)
library(dplyr)
library(ggplot2)

load("local")

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      radioButtons(
        inputId = "peak",
        label = "Devicce:",
        choices = c("All types", "Desktop", "Mobile"))),
    mainPanel(plotOutput(outputId = "heatmap", height = "70vh"))))

server <- function(input, output){
  dataInput <- reactive({
    switch(input$peak,  
           "All types" = local %>% filter(device == "all"),
           "Desktop" = local %>% filter(device == "desktop"),
           "Mobile" = local %>% filter(device == "mobile"))
  })
  
  output$heatmap <- renderPlot({
    ggplot(dataInput(), aes(localdate, localhour2)) + 
      geom_tile(aes(fill = n), colour = "white") + 
      scale_fill_gradient(low = "#f3da4c", high = "blue", breaks = seq(0, 13, 2)) +
      scale_x_date(breaks = seq(as.Date("2016-01-26 UTC"), as.Date("2016-04-02 UTC"), "7 days"), date_labels = "%b %d") + 
      scale_y_continuous(breaks = seq(0, 23, 1)) +
      labs(x = "Date", y = "Hour", fill = "Freq of Online Access") +
      ggtitle("Peak Hours of Studying Online (Local Time)") 
  })
}

shinyApp(ui = ui, server = server)
```


