---
title: "Data Transformation with Base R and dplyr"
author: "Yun Dai"
date: "04/2019"
output: 
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    toc_float: true
    css: contents.css
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
load("/Volumes/GoogleDrive/My Drive/websites/new web/data/r/sample/sample")
subsample <- data.frame(sample)
```


******
In this post, we will go through the tasks we most frequently meet with in data transformation. We will take a very practical approach, at times even seemingly trivial, leaving the conceptual part aside for a moment. 

We are going to show examples along the way with the aid of package `dplyr` as well as base R functions. `dplyr` provides fast and intuitive ways to manipulate date frames. [This article](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html) offers a quick start of `dplyr` on data frame manipulation essentials.

Before we start, let's load the package `dplyr`. 
```{r message=FALSE, warning=FALSE}
library(dplyr)
```

******
A side note on `dplyr` pipe operator `%>%` before we proceed. 

The `dplyr` pipe operator `%>%` originally comes from package `magrittr`. In `dplyr`, `%>%` chains functions together, passing the output of the former function to the input of the next function. This way we don't have to nest a lot of functions, which also makes the code more readable.

For instance, in the example below, step by step each new operation is performed on the output from the previous one. We first keep the rows where the variable `resource` has no missing cases; we then group the data by `date` and `resource` on the filtered subset; we next add a new column to count observations of the newly formed group; following that we rename the new variable counting the observations, and so forth. 
```{r eval=FALSE}
session %>% 
  filter(!is.na(resource)) %>%
  group_by(date, resource) %>%
  add_tally(interval) %>%
  rename(sum_day = n)
```

******
## Recoding variables

******
### variable transformation
Variable transformation is the replacement of a variable by a new value or a function of that variable. The purposes can be multiple: convenience, reducing skewness, producing linear or additive relationships, and more. Standardisation and logarithmic transformation are common types that are familiar to us. 

Below we are not discussing why we are performing a task, but simply showing how with a few examples for specific tasks.

******
### `<-` to assign new values
One way we often use to recode variables is to use the assignment sign `<-`.

Let's use the sample dataset `storms` from `dplyr` to show how we may recode the missing values.
```{r}
head(dplyr::storms)
```

As we can see, there are many missing values in `storms$ts_diameter`.
```{r}
summary(storms$ts_diameter)
```

We decide that we want to recode all `NA`s to 9999.
```{r}
storm <- storms

storm$ts_diameter[is.na(storm$ts_diameter)] <- 9999
summary(storm$ts_diameter)
```

We can of couse recode the values 9999 back to `NA`.
```{r}
storm$ts_diameter[storm$ts_diameter == 9999] <- NA
summary(storm$ts_diameter)
```

******
### `recode()` to replace values
`recode()` replaces numeric values based on their position, and replaces character values by their name. `.default` argument, if supplied, gives all cases not matched a new value. `.missing` argument replaces the missing values by the specified new value.

Recoding character values:
```{r}
iris$species <- recode(iris$Species, setosa = "a",  versicolor = "b", virginica = "c")
table(iris$species)
```

Recoding numeric values:
```{r}
mtcars$cyl2 <- recode(mtcars$cyl, `4` = 1, `6` = 2, `8` = 3)
table(mtcars$cyl2)
```

******
### `case_when()` to recode on subsets
`case_when()` allows us to vectorize multiple `if` and `else if` statements. 

For instance, below we reorganized the majors by grouping them to a higher category.
```{r eval=FALSE}
survey$new_major <- case_when(
  survey$major %in% c("Biology", "Chemistry", "Physics", "Neural Science") ~ "Science",
  survey$major %in% c("Computer Systems Engineering", "Electrical and Systems Engineering") ~ "CS & Engineering",
  survey$major %in% c("Business and Marketing", "Business and Finance", "Economics") ~ "Business, Finance & Economics",
  survey$major %in% "Mathematics" ~ "Mathematics",
  survey$major %in% NA ~ "Undefined"
)
```

******
### `mutate()` to add new variables at the same time
`mutate()` adds new variables while recoding them.

```{r}
mt <- mtcars %>% mutate(newvar = mpg * 2)
head(mt)
```

******
`mutate()` can operate on multiple variables.
```{r}
mt <- mtcars %>% mutate(newvar1 = mpg * 2, newvar2 = cumsum(mpg))
head(mt)
```

******
### `mutate(ifelse())` to recode with conditions
`mutate()` can be used with functions. For instance, `mutate()` with `ifelse()` recodes variables conditionally.

Let's say we want to create a new column, copying cases whose values are smaller than 3.
```{r eval=FALSE}
mutate(rank2 = ifelse(rank <= 3, rank, NA))
```


In another example, let's say we have a dataset `subsample`. 
```{r}
subsample
```

When processing the data, we want to combine the information from two columns. The rule is that we only need the non-missing case from the two columns.
```{r}
subsample %>% mutate(
  newvar = ifelse(is.na(Q1), Q2, NA),
  newvar = ifelse(is.na(Q2), Q1, newvar)
)
```


******
### `mutate(case_when())`
`case_when()` can be nested within `mutate()`.
```{r eval=FALSE}
mutate(rank2 = case_when(
    rank <= 3 ~ rank,
    rank > 3 & row_number() %% 2 == 0 & expd == 1 ~ 4,
    rank > 3 & row_number() %% 2 == 1 & expd == 1 ~ 5))
```


******
## Adding / dropping columns and rows
There are several ways to add and drop columns and rows in a data frame. First, let's look at the base R methods using the sample dataset `mtcars`.

```{r}
head(mtcars)
```

Adding a column:
```{r}
mt <- mtcars
mt$new_col <- 2 * mt$cyl
head(mt)
```

Removing a column:
```{r}
mt$new_col <- NULL
head(mt)
```

Removing a row:
```{r}
mt <- mtcars[-(1:5), ]
head(mt)
```

******
### `select()`
`dplyr` offers several functions that handle adding and dropping columns and rows efficiently.

`select()` keeps the selected variables.
```{r}
mt <- mtcars %>% select(mpg, cyl) 
head(mt)
```

We can rename the variables while we select the variables.
```{r}
mt <- mtcars %>% select(mpg2 = mpg, cyl2 = cyl) 
head(mt)
```


******
### `filter()`
`filter()` filters rows, keeping rows where given conditions are true.

Here we use the sample dataset `iris`.
```{r}
head(iris)
```

Below we filter the rows to get a subset where `Species` is all `versicolor` and `Petal.Width` is smaller than 1.2
```{r}
filter(iris, Species == "versicolor" & Petal.Width < 1.2)
```

This would be the same as
```{r}
iris %>% filter(Species == "versicolor" & Petal.Width < 1.2)
```


******
## Renaming variables

******
### `names()`
In base R, we can rename variables using function `names()`.
```{r}
mt <- mtcars
names(mt)[1] <- "var"
names(mt)
```

Renaming more than one variable:
```{r}
names(mt)[1:3] <- c("a", "b", "c")
names(mt)
```

Renaming all:
```{r}
names(mt)[1:11] <- letters[1:11]
names(mt)
```


******
### `rename()`
We can also use `rename()` from `dplyr`, which renames the variables without dropping variales. If we use `select()` to rename variables, the unmentioned variables will be dropped.
```{r}
mt <- mtcars %>% rename(mpg2 = mpg, cyl2 = cyl) 
head(mt)
```

Compare it with how `select()` works:
```{r}
mt <- mtcars %>% select(mpg2 = mpg, cyl2 = cyl) 
head(mt)
```



******
## Arranging rows
`arrange()` from package `dplyr` sorts variables in ascending order, and `desc()` sorts variables in descending order.

This example below sorts `mtcars` by variable `mpg` in ascending order.
```{r}
df <- arrange(mtcars, mpg)
head(df)
```

`arrange()` can sort multiple variables in the order specified. The example below sorts `df` first by `mpg`, then `cyl`, and finally `disp`.
```{r}
df <- arrange(mtcars, mpg, cyl, disp)
head(df)
```

This example sorts `df` first by `mpg` and `cyl` in ascending order, and then `disp` in descending order.
```{r}
df <- arrange(mtcars, mpg, cyl, -disp)
head(df)
```

Same as above, except that we use function `desc()` to sort rows in descending order.
```{r}
df <- arrange(mtcars, mpg, cyl, desc(disp))
head(df)
```

******
## Reordering columns
We can move columns with `select(var, everything())`.
```{r}
new_col <- mtcars %>% select(carb, gear, everything())
head(new_col)
```

Or move a column the other way:
```{r}
new_col <- mtcars %>% select(-mpg, everything())
head(new_col)
```


******
## Grouping
`group_by()` from `dplyr` groups a data frame by the specified variable. `ungroup()` removes the grouping.
```{r}
storm <- dplyr::storms %>% group_by(status)
head(storm)
```


******
`group_by()` is most useful to use with `summarise()`, `filter()`, `mutate()`, and `arrange()`.
```{r message=FALSE, warning=FALSE}
storm <- dplyr::storms %>% filter(!is.na(ts_diameter)) %>% group_by(year) %>% summarise(mean = mean(ts_diameter))
head(storm)
```


******
## Removing duplicates
`distinct()` from `dplyr` selects unique rows from a data frame.

```{r eval=FALSE}
distinct(mtcars)
```

In the example below, duplicates are removed and all other variables are kept where `cyl` has no duplicates. Only the first cases of duplicated `cyl` have been kept.

```{r}
distinct(mtcars, cyl, .keep_all = TRUE)
```

Without argument `.keep_all`, the other variables will be dropped. 
```{r}
distinct(mtcars, cyl)
```

`distinct()` allows for multiple variables to be included. 
```{r}
distinct(mtcars, cyl, gear, .keep_all = TRUE)
```

******
`unique()` from base R can also extract unique elements, but `distinct()` offers more flexibility.