---
title: "Duplicates"
author: "Yun Dai"
date: "05/2020"
output: 
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    toc_float: true
    css: contents.css
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
rm(list = ls())
library(readr)
location <- read_csv("/Volumes/GoogleDrive/My Drive/websites/new web/data/r/sample/faculty_loc.csv")
```

***
In this post, we discuss a few tricks for working with duplicates, including how to find and mark duplicates, find positions of duplicates, and remove duplicates.

The sample dataset we will be using is shown below. In this dataset, `Name` and `Location` both contain duplicates.

```{r echo=FALSE}
location
```

 
***
## Finding duplicates
Using `duplicated()`, we will get a logical vector indicating which values are duplicates. By default, the first value will be marked as the unique value (`FALSE`) and the rest as duplicates (`TRUE`).

```{r}
duplicated(location$Name)
```

***
We may create a new column to mark the duplicates.
```{r}
location$dup <- duplicated(location$Name)
```

```{r echo=FALSE}
location
```

***
In cases where we want to mark all identical elements as "duplicates", we can specify the `fromLast` argument to be `TRUE`. When `fromLast` is `TRUE`, duplication would be considered from the reverse side.
```{r}
location$dup2 <- duplicated(location$Name) | duplicated(location$Name, fromLast=TRUE)
```

```{r echo=FALSE}
location
```

***
## Finding positions of duplicates
`which()` gives us the TRUE indices of a logical object.

Using `which()` with `duplicated()`, we get the positions of the duplicates.
```{r}
which(duplicated(location$Name))
```

***
## Removing duplicates
`dplyr::distinct()` keeps the unique values of selected columns.

If we do not specify `.keep_all = TRUE`, only variable `Name` will be kept and the others will be dropped.
```{r message=FALSE,warning=FALSE}
library(dplyr)

location %>% distinct(Name, .keep_all = TRUE)
```

***
We can include multiple variables in `distinct()` so that the combination of the group would be unique.

```{r}
location %>% distinct(Name, Location, .keep_all = TRUE)
```

