---
title: "Flow Control"
author: "Yun Dai"
date: "07/2021"
output: 
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    toc_float: true
    css: contents.css
---

***
There are times when we would like to execute some code if a *condition* is fulfilled. 

This post discusses flow control. We will see conditional statements.

***
## `if`
The simplest form of flow control is `if` conditional statements. `if` takes a logical value (a logical vector of length one). If the condition is `TRUE`, the statement gets executed. If the condition is `FALSE`, `if` invisibly returns `NULL`.

***
*How it works:*

If we only have one statement:
```
if(condition) statement
```

If we have more than one statement, we need to use a pair of curly braces to wrap the statements.
```
if(condition) {
  statement
}
```

***
*Example:*
```{r}
x <- sample(1:15, 1) 
x

if (x > 5) {
  print(paste(x, "is larger than 5."))
}
```

Or:
```{r}
x <- sample(1:10, 1) 
x

if (x > 5) print(paste(x, "is larger than 5."))
```


***
## `if` ... `else if`
If there are other conditions to evaluate, include `else if`. `else if` is essentially saying "if the previous condition is not true, then try this condition". 

***
*How it works:*
```
if(condition) {
  statement
} else if (condition) {
  statement
} 
```

***
*Example:*
```{r}
x <- sample(1:15, 1)
x

if (x > 5) {
  print(paste(x, "is larger than 5."))
} else if (x < 5) {
  print(paste(x, "is smaller than 5."))
}
```

***     
We can have more than one `else if`.
```
if(condition) {
  statement
} else if (condition) {
  statement
} else if (condition) {
  statement
}
```

***
## `if` ... `else if` ... `else`
`else` catches anything that is not caught by the preceding conditions. The `else` statement is only evaluated if the condition is `FALSE`.

***
*How it works:*

```
if(condition) {
  statement
} else if (condition) {
  statement
} else {
  statement
}
```

***
*Example 1:*
```{r}
x <- sample(1:15, 1)
x

if (x > 5) {
  print(paste(x, "is larger than 5."))
} else if (x == 5) {
  print(paste(x, "is equal to 5."))
} else {
  print(paste(x, "is smaller than 5."))
}
```

Note: The `else` statement must occur on the same line as the closing curly brace from the `if` clause. 

***
*Example 2:*

Write an R program to convert month name to a number of days. For instance, if we input "February", then the output will be "No. of days: 28 or 29 days". Hint: Use `readline()`, which reads a line from the terminal in interactive use.

```{r eval=FALSE}
month <- as.character(readline("Input a month name: "))
month_30 <- c("April", "June", "September", "November")
month_31 <- c("January", "March", "May", "July", "August", "October", "December")

if (month %in% month_30){
  print("No. of days: 30 days") 
} else if (month %in% month_31){
  print("No. of days: 31 days")
} else if (month == "February") {
  print("No. of days: 28 or 29 days")
} else {
  print("Incorrect input")
}
```


***
## `if` ... `else`
We can have an `else` statement without the `else if`.

***
*How it works:*
```
if(condition) {
  true statement
}
else {
  false statement
}
```

***
*Example:*
```{r}
x <- sample(1:15, 1)
x

if (x > 5) {
  print(paste(x, "is larger than 5."))
} else {
  print(paste(x, "is not larger than 5."))
}
```


***
## Nested `if`
We can have `if` statements inside `if` statements. This is called nested `if` statements.

***
*Example 1:*
```{r}
x <- sample(1:15, 1)
x

if (x > 5) {
  print(paste(x, "is larger than 5,"))
  if (x > 10) {
    print(paste("and 10."))
  } else {
    print("but not larger than 10.")
  }
} else {
  print(paste(x, "is not larger than 5"))
}
```

***
*Example 2:*

Write an R program to find the median of three integers. Do not use the built-in function `median().`

```{r eval=FALSE}
a <- as.integer(readline("Input the 1st number: "))
b <- as.integer(readline("Input the 2nd number: "))
c <- as.integer(readline("Input the 3rd number: "))

if (a > b){
  if (b > c) { print(b) }
  else if (c > a) { print(a) } 
  else { print(c) }
} else {
  if (a > c){ print(a) }
  else if (c > b){ print(b) } 
  else{ print(c) }
} 
```


***
## Vectorized `ifelse()`
Remember that `if` only takes a logical vector of length *one*. What if we have multiple logical values? 

If we pass a logical vector with a length of more than one, e.g. `if(c(TRUE, FALSE))`, only the first one will be used. 

There is a vectorized version of the `if ... else` construct, the `ifelse()` function. 

***
*How it works:*
```
ifelse(condition, yes, no)
```

`ifelse()` takes three arguments. The first is a logical vector of conditions. The second contains values that are returned when the first vector is `TRUE`. The third contains values that are returned when the first vector is `FALSE`.

It returns a vector of the same length as `condition`, with elements `yes[i]` if `condition[i]` is `TRUE`, otherwise `no[i]`. `yes` and `no` are recycled where necessary.

***
*Example:*
```{r}
x <- sample(1:15, 1)
x

yes <- paste(x, "is larger than 5")
no <- paste(x, "is not larger than 5")
ifelse(x > 5, yes, no)
```
