---
title: "Functions"
author: "Yun Dai"
date: "07/2021"
output: 
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    toc_float: true
    css: contents.css
---

******
Control statements are most often used in connection with functions. As we touched upon briefly last week, a function is an object in R that takes some input objects (arguments) and returns an output object. 

Think about functions as tools that we can use to do things. We have been working with built-in functions so far. We can also define our own functions in R, assign them a name, and then call them just like the built-in functions. 

***
## Creating and calling functions
To create a function, use the `function` keyword, followed by a list of arguments and the function body. 

***
*How it works:*

If the function body is a single expression:
```
function(argument1, ...., argumentN) expression
```

If there are a series of expressions, curly braces should be used around the function body:

```
function(argument1, ...., argumentN) {
  expression1
  .
  .
  .
  expressionrM
}
```

To call a function, use the function name followed by parenthesis.

***
*Example:*
```{r}
my_function <- function(x,y,z) {c(x + 1, y + 1, z + 1)}
my_function(1,2,3)
```

***
A useful trick: We can type the name of a function to see the code that runs when we call it.
```{r}
my_function
```


***
## Arguments 
When we call a function in R, there are different ways to specify the arguments. 

When arguments are given in the named form `name = object`, they may be given in any order. 
 
If we don’t name the arguments, then R will match them based on position. 

Besides, we may also specify named arguments after the positional arguments.

```{r}
my_function(1,2,3)
my_function(x = 1, y = 2, z = 3)
my_function(1, 2, z = 3)
```

However, it’s a good practice to use exact argument names. It removes ambiguity and makes our code more readable. 

***
### R functions can take default parameter values. 
If we don’t pass anything to the function, it uses the default value. 

```{r}
my_function <- function(x = 1, y = 2, z = 3) {c(x + 1, y + 1, z + 1)}
my_function()
my_function(x = 3, y = 4)
```


***
*Example 1:*

Write an R program to calculate a dog's age in dog's years. For the first two years, a dog year is equal to 10.5 human years. After that, each dog year equals 4 human years. When a dog's age in human years is 15, what is the dog's age in dog's years?

```{r}
age <- function(human){
  if (human <= 2){
    dog <- human*10.5
  } else {
    dog <- 2*10.5 + (human-2)*4
  }
  print(dog)
}
age(15)
```


***
*Example 2:*

Write an R function that calculates the number of uppercase letters and lowercase letters in a string. Hint: Use `?"regular expression"`.

```{r}
function_letters <- function(string) {
  string <- strsplit(string, split = "")[[1]]
  c1 <- 0
  c2 <- 0
  for (i in string){
    if (grepl("[A-Z]", i)){
      c1 <- c1 + 1  
    } else if (grepl("[a-z]", i)){
      c2 <- c2 + 1
    }
  }
  print(paste("upper case: ", c1))
  print(paste("lower case: ", c2)) 
}

string1 <- "Mr. and Mrs. Dursley of number four, Privet Drive, were proud to say that they were perfectly normal, thank you very much."
string2 <- "One hot spring evening, just as the sun was going down, two men appeared at Patriarch's Ponds."

function_letters(string1)
function_letters(string2)
```


***
## Returning values 
R returns the last evaluated expression in the function. We may also use the `return()` function to specify the value returned by the function. But it is common to omit the return statement.

```{r}
my_function <- function(x,y,z) {return(c(x + 1, y + 1, z + 1))}
my_function(1,2,3)
```


***
### `print()` or `return()`
`print()` shows the human user a string representing what is going on inside the computer. `return()` gives back a value. 

`print()` is very useful for understanding how a program works and can be used in debugging to check various values in a program without interrupting the program.

`return()` is the main way that a function returns a value. All functions will return a value, and if there is no return statement, it will return nothing. The value that is returned by a function can then be further used as an argument passed to another function, stored as a variable, or just printed for the benefit of the human user.

***
*Example:*  

`return()`:
```{r}
tri <- function(k) {
  if (k > 0) {
    result <- k + tri(k - 1)
    return(result)
  } else {
    result <- 0
    return(result)
  }
}
tri(6)
```

***
`print()`:
```{r}
tri <- function(k) {
  if (k > 0) {
    result <- k + tri(k - 1)
    print(result)
  } else {
    result <- 0
    print(result)
  }
}
tri(6)
```


***
## `stop()`
`stop()` stops execution of the current expression and executes an error action.

```{r}
deposit <- function(amount) {
  total <- 100
  if(amount <= 0)
  stop("Deposits must be positive!\n")
  total <- total + amount
  print(paste(amount, "deposited. Your balance is", total))
}

deposit(6)
```

If we try to deposit -6 into our account by `deposit(-6)`, the function will return the error message `Error in deposit(-6) : Deposits must be positive!`.

***
## Recursive functions
In R, a function can call itself. We can loop through data to reach a result. Let's first see an example; then we'll explain it, and we will discuss three more examples (problem sets).

***
*Example:*  

```{r}
tri <- function(k) {
  if (k > 0) {
    result <- k + tri(k - 1)
    return(result)
  } else {
    result <- 0
    return(result)
  }
}
tri(6)

## Code from https://www.w3schools.com/r/r_functions.asp
```


***
*Notes:*
(1) Recursion

k      result             
-----  ------- 
6      6+tri(5)             
5      5+tri(4)              
4      4+tri(3)          
3      3+tri(2)           
2      2+tri(1)              
1      1+tri(0)
0      0


(2) Returned values

result    output           
------    --------------     
tri(0)    0          
1+tri(0)  1            
2+tri(1)  2+1=3         
3+tri(2)  3+2+1=6        
4+tri(3)  4+3+2+1=10             
5+tri(4)  5+4+3+2+1=15
6+tri(5)  6+5+4+3+2+1=21


***
## Nested functions
We can call a function within another function.

We'll also see passing functions to other functions with the `apply` family function in data frame manipulation, which we discuss in data manipulation.

```
stdError <- function(x) {sqrt(var(x)/length(x))}
tapply(data, label, stdError)
```

***
## Scope of variables
A variable’s scope is the set of places from which you can see the variable. R will try to find variables in the current environment, and if it doesn’t find them it will look in the parent environment, and then that environment’s parent, and so on until it reaches the global environment. 

(1) 
```{r}
y <- 10
f <- function(x) {
  y <- 2
  y*2 + g(x)
}

g <- function(x) {
  x * y
}

f(3)
```

***
(2)
```{r}
f <- function(x) {
  y <- 2
  g <- function(x) {
    x * y
  }
  y*2 + g(x)
}

f(3)
```


