---
title: "global"
author: "Yun Dai"
output: 
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    toc_float: true
---

```{r echo=FALSE}
knitr::opts_knit$set(root.dir = "/Volumes/GoogleDrive/My Drive/websites/new web/archived/r sample files/globalR") 
```

******
This page includes a small portion of the global.R script for running the App, but should be representative of many scenarios discussed in the series of posts.

******
Load all libraries that will be needed.
```{r message=FALSE}
library(readxl)
library(dplyr)
library(stringr)
library(data.table)
library(ggplot2)
```

******
Load data. Define the color palette that will be used later in plotting.
```{r message=FALSE}
load("survey")

# color palette
palette <- list(purple = c("#351F39", "#351C4D", "#6c1f55", "#765285", "#8a6899" ),
                turquoise = c("#709FB0", "#849974", "#A0C1B8"),
                golden = c("#D1A827", "#f3da4c"))
```

******
### Summary table
```{r message=FALSE}
set <- survey[c("Q7.3", "Q7.2", "Q7.4", "country")]
tb <- replicate(length(set), data.frame(), simplify=FALSE)
for (m in 1:length(set)){
  tb[m] <- list(data.frame(table(set[m])))
  tb[[m]]$`%` <- round(tb[[m]]$Freq/320 *100, 2)
  names(tb[[m]]) <- c("", "#", "%")
  tb[[m]] <- tb[[m]][order(tb[[m]]$`#`,decreasing = T),]
}
## replicate() creates a list of 4 df; each df for one summary table
## tb[m] <- list() adds one df by another
## tb[[m]] extracting df
## tb[[m]][order(tb[[m]]$`#`,decreasing = T),] orders rows by freq

for (i in 1:4){
  colnames(tb[[i]])[1] <- c("Student Status", "Country / Region", "Major", "country")[i]
}

tb[[2]][,2:3] <- NULL 

tb
```

******
### Service use
Define groups for questions of multiple choices.
```{r message=FALSE}
g <- list(status = c("Freshman", "Sophomore", "Junior", "Senior", "Study Away", "Graduate/Joint Program") ,
          country = c("China", "U.S.", "Other"),
          major = c("Science, CS, Engineering, Mathematics", 
                    "Business, Finance, Economics", 
                    "Humanities, Social Sciences", 
                    "Data Science, Interactive Media Business", 
                    "Interactive Media Arts"))
```

******
#### Visiting Library 
```{r message=FALSE}
## Q2.2: My top reasons for visiting Library

Reason <- c("Work on a class assignment/paper", 
            "Watch video or listen audio", "Use specialized databases \\(e.g. Bloomberg, Wind\\)",
            "Use a library computer",  "Use a group study room", "Print, photocopy, scan",  "Other", 
            "Meet up with friends" , "Hang out between classes", 
            "Get readings from Course Reserve", "Get help from a librarian", "Find a quiet place to study",
            "Borrow books and materials", "Attend a library workshop")

r <- data.frame(survey$status, survey$country, survey$major2)
for (m in 1: length(Reason)){
  r[,Reason[m]] <- str_extract(survey$Q2.2, Reason[m])
}

dtset <- replicate(3, data.frame(Reason), simplify = FALSE)
for (i in 1:3){
  levels(dtset[[i]]$Reason)[levels(dtset[[i]]$Reason) == "Use specialized databases \\(e.g. Bloomberg, Wind\\)"] <- "Use specialized databases (e.g. Bloomberg, Wind)"
}

for (m in 1:length(dtset)){
  for (n in 1:length(g[[m]])){
    subgroup <- r[r[,m] == g[[m]][n], (length(g)+1):(length(Reason) + length(g))]
    dtset[[m]][,g[[m]][n]] <- apply(subgroup, 2, function(x) length(which(!is.na(x))))
  }
  dtset[[m]]$Total<- rowSums(dtset[[m]][,2:n], na.rm = TRUE, dims = 1)
  dtset[[m]] <- dtset[[m]][order(dtset[[m]]$Total,decreasing = T),]
}

## m denotes how many groups we have. We have 3 groups: country/status/major
## n denotes how many elements we have in each group. e.g. 3 elements in the group "country" (China, U.S., Other)

Reason2 <- c("Work on a class assignment/paper", "Watch video or listen audio", 
             "Use specialized databases (e.g. Bloomberg, Wind)",
             "Use a library computer",  "Use a group study room", "Print, photocopy, scan",  "Other", 
             "Meet up with friends", "Hang out between classes", "Get readings from Course Reserve", 
             "Get help from a librarian", "Find a quiet place to study",
             "Borrow books and materials", "Attend a library workshop")
l <- data.frame(Reason = Reason2, Level = c("High","Low","Low","Medium","Medium","High","Low","Medium","Low","Medium","Low","High","High","Low"))

for (i in 1:3){
  dtset[[i]] <- dtset[[i]] %>% left_join(l, by = "Reason")
}
## labeling; for grouping in graphing

dtset
```

******
### Space and study habit
#### Space Preference: Most Days - ranking
```{r message=FALSE}
survey <- survey %>% mutate(
  mostdays_crowded = ifelse(is.na(Q2.5_0_15_RANK), Q2.5_0_1_RANK, NA),
  mostdays_crowded = ifelse(is.na(Q2.5_0_1_RANK), Q2.5_0_15_RANK, mostdays_crowded),
  mostdays_modpop = ifelse(is.na(Q2.5_0_16_RANK), Q2.5_0_9_RANK, NA),
  mostdays_modpop = ifelse(is.na(Q2.5_0_9_RANK), Q2.5_0_16_RANK, mostdays_modpop),
  mostdays_noisy = ifelse(is.na(Q2.5_0_17_RANK), Q2.5_0_3_RANK, NA),
  mostdays_noisy = ifelse(is.na(Q2.5_0_3_RANK), Q2.5_0_17_RANK, mostdays_noisy),
  mostdays_quiet = ifelse(is.na(Q2.5_0_18_RANK), Q2.5_0_4_RANK, NA),
  mostdays_quiet = ifelse(is.na(Q2.5_0_4_RANK), Q2.5_0_18_RANK, mostdays_quiet),
  mostdays_silent = ifelse(is.na(Q2.5_0_19_RANK), Q2.5_0_5_RANK, NA),
  mostdays_silent = ifelse(is.na(Q2.5_0_5_RANK), Q2.5_0_19_RANK, mostdays_silent),
  mostdays_relaxed = ifelse(is.na(Q2.5_0_20_RANK), Q2.5_0_6_RANK, NA),
  mostdays_relaxed = ifelse(is.na(Q2.5_0_6_RANK), Q2.5_0_20_RANK, mostdays_relaxed),
  mostdays_focused = ifelse(is.na(Q2.5_0_21_RANK), Q2.5_0_7_RANK, NA),
  mostdays_focused = ifelse(is.na(Q2.5_0_7_RANK), Q2.5_0_21_RANK, mostdays_focused)
)
## keep only if one of the pair is missing and the other non-missing
## this is due to how the Qualtrics survey was designed; not important for our purpose here

days_status <- 
  data.frame(survey %>% group_by(status) %>% summarise(mostdays_crowded = round(mean(mostdays_crowded, na.rm =TRUE),2))) %>% 
  left_join(data.frame(survey %>% group_by(status) %>% summarise(mostdays_modpop = round(mean(mostdays_modpop, na.rm =TRUE),2)))) %>%
  left_join(data.frame(survey %>% group_by(status) %>% summarise(mostdays_noisy = round(mean(mostdays_noisy, na.rm =TRUE),2)))) %>%
  left_join(data.frame(survey %>% group_by(status) %>% summarise(mostdays_quiet = round(mean(mostdays_quiet, na.rm =TRUE),2)))) %>%
  left_join(data.frame(survey %>% group_by(status) %>% summarise(mostdays_silent = round(mean(mostdays_silent, na.rm =TRUE),2)))) %>%
  left_join(data.frame(survey %>% group_by(status) %>% summarise(mostdays_relaxed = round(mean(mostdays_relaxed, na.rm =TRUE),2)))) %>%
  left_join(data.frame(survey %>% group_by(status) %>% summarise(mostdays_focused = round(mean(mostdays_focused, na.rm =TRUE),2)))) %>%
  rename(`crowded` = mostdays_crowded, `moderately populated` = mostdays_modpop, 
         `noisy` = mostdays_noisy, `quiet (occasional whispers)` = mostdays_quiet, 
         `(close to) silent` = mostdays_silent, `relaxed, cozy or homey` = mostdays_relaxed,
         `focused` = mostdays_focused, `G` = status) %>% 
  melt()

## same for creating the country and major matrices 
```

******
#### Space Preference: Most Days - multiple options
```{r message=FALSE}
## Q2.5_0_GROUP: most days

Place <- c("crowded", "moderately populated", "noisy", "quiet \\(occasional whispers\\)", "\\(close to\\) silent", "relaxed, cozy or homey", "focused")

p1 <- data.frame(survey$status, survey$country, survey$major2)
for (m in 1:length(Place)){
  p1[,Place[m]] <- str_extract(survey$Q2.5_0_GROUP, Place[m])
}

dtset4 <- replicate(3, data.frame(Place), simplify = FALSE)
for (m in 1:length(dtset4)){
  for (n in 1:length(g[[m]])){
    subgroup4 <- p1[p1[,m] == g[[m]][n], (length(g)+1):(length(Place) + length(g))]
    dtset4[[m]][,g[[m]][n]] <- apply(subgroup4, 2, function(x) length(which(!is.na(x))))
  }
  dtset4[[m]]$Total<- rowSums(dtset4[[m]][,2:n], na.rm = TRUE, dims = 1)
}

for (i in 1:3){
  levels(dtset4[[i]]$Place)[levels(dtset4[[i]]$Place) == "quiet \\(occasional whispers\\)"] <- "quiet (occasional whispers)"
  levels(dtset4[[i]]$Place)[levels(dtset4[[i]]$Place) == "\\(close to\\) silent"] <- "(close to) silent"
}

dtset4
```

```{r message=FALSE}
## grand matrix - most days
days_status <- days_status %>% 
  left_join(rename(melt(dtset4[[1]]), `G` = variable, `variable` = Place, `count` = value ), by = c("G", "variable")) 
days_status$G <- factor(days_status$G, levels = c("Freshman", "Sophomore", "Junior", "Senior", 
                                                  "Graduate/Joint Program", "Study Away"))

## same for creating the country and major matrices 
```


******
#### Space Preference: Student Submissions
```{r message=FALSE}
lib <- unlist(strsplit(survey$Q2.3.y, ","))
lib <- str_trim(lib, side = "both")
lib <- lib[!lib %in% c("NULL")]
lib <- data.frame(table(lib)) %>% arrange(-Freq)

color1 <- rep(c(palette[[1]][4], palette[[2]][1], palette[[3]][1], palette[[2]][2], palette[[2]][3], palette[[3]][2]),
              length.out=nrow(lib))

## for later use in wordcloud2 package

lib
```

