---
title: "Loops"
author: "Yun Dai"
date: "07/2021"
output: 
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    toc_float: true
    css: contents.css
---

***
Loops execute a block of code as long as a specified condition is reached.

We discuss two kinds of loops in R: `for` loops and `while` loops.

***
## `for` loops
A `for` loop is used for iterating over items in a vector (or an array or a list). It is useful if we know in advance the set of values that we want to iterate over. 

***
*How it works:*
```
for (variable in vector) expression
```

*variable* is the loop variable. For each *variable* in *vector*, *expression* is called once; the value of *variable* is then updated. *vector* can be a vector, array, list, etc.

*expression* is often a grouped expression. In that case, when we have multiple expressions, the expressions must be surrounded by curly braces.

```
for (variable in vector) {
  expression 1
  expression 2
  ...
}
```

***
It's conventional to use very short variable names like `i` or `j` to iterate over a vector.

```{r}
for (i in seq(from = 5, to = 25, by = 5)) print(i)
```


*** 
In loops, we must use the `print()` function if we want to output a result. 

```{r}
for (i in 1:5) {
  i
}
```


***
*Example 1:*

Append each value in ` c(0,1,2,3,4,5)` to an empty vector. 

```{r}
values <- c(0,1,2,3,4,5)

vec <- numeric(0)
#vec <- vector("numeric", 0) returns a numeric vector of length 0
#vec <- c() returns a NULL object
#vec <- vector() returns a logical vector of length 0

for (i in seq_along(values)) {
  vec <- c(vec, values[i])
}
print(vec)
```


***   
*Notes:*

(1) `seq_along(values)` generates a sequence `1,2,3,4,5,6`.

(2) `vec <- c(vec, values[i])` flattens vectors and combines them into one single vector. When the arguments to `c()` are themselves vectors, it flattens them and combines them into one single vector.

(3) `for (i in seq_along(values))` iterates over each item in `values`.    

vec                 i     values[i]    
------------------  ----  ----------
empty                     
0                   1     0    
0,1                 2     1
0,1,2               3     2
0,1,2,3             4     3
0,1,2,3,4           5     4
0,1,2,3,4,5         6     5

    
(4) We can substitute `i in seq_along(values)` with `i in 1:length(values)` in this case. 
```{r}
vec <- c()
for (i in 1:length(values)) {
  vec[i] <- values[i]
}
print(vec)
```

There are times that `seq_along(x)` and `1:length(x)` do not return the same results; that's when `x` is empty. 

```{r}
x <- vector("numeric")
seq_along(x)
1:length(x)
```

   
We will be seeing the trick of appending values to an empty vector (or other objects) frequently in writing loops.

***
*Example 2:*

Get the first 10 [Fibonacci numbers](https://en.wikipedia.org/wiki/Fibonacci_number). The Fibonacci numbers form a sequence, such that each number is the sum of the two preceding ones, starting from 0 and 1. 

```{r}
f <- numeric(10)
f[1] <- 0
f[2] <- 1
for (i in 3:10) {
  f[i] <- f[i - 2] + f[i - 1]
}  
print(f)
```


***  
*Notes:*

f                         i      f[i]    f[i-2]   f[i-1]
------------------------  -----  ------  -------  -------
0                         1      0
0,1                       2      1    
0,1,1                     3      1       0        1
0,1,1,2                   4      2       1        1
0,1,1,2,3                 5      3       1        2
0,1,1,2,3,5               6      5       2        3
0,1,1,2,3,5,8             7      8       3        5
0,1,1,2,3,5,8,13          8      13      5        8
0,1,1,2,3,5,8,13,21       9      21      8        13
0,1,1,2,3,5,8,13,21,34    10     34      13       21


***
*Example 3:*

Count how many 2 occurs in the vector `rep(1:4, each = 2, times = 3)`. 
```{r}
vec <- rep(1:3, each = 2, times = 2)
c <- 0
for (i in seq_along(vec)){
  if (vec[i] == 2){
    c <- c + 1
  }
}

print(c)
```

   
***
*Notes:*

i      vec[i]  TRUE     c
-----  ------  -------  ------
1      1       0        0
2      1       0        0
3      2       1        1
4      2       1        2
5      3       0        2
6      3       0        2
7      1       0        2
8      1       0        2
9      2       1        3
10     2       1        4
11     3       0        4
12     3       0        4


In real life, you should just use the vectorized `sum()`. What we did was just to get familiar with how a `for` loop works. 
```{r}
vec <- rep(1:4, each = 2, times = 3)
sum(vec == 2)
```

***
*Note:* Wherever vectorized operations are possible, we should avoid `for` loops. `for` loops in R run much slower than their vectorized equivalents. 

***
*Example 4:*

Print the numbers from 1 to 100. Print "Fizz" for multiples of 3, print "Buzz" for multiples of 5, and print "FizzBuzz" for multiples of both 3 and 5. Hint: Use the modulus operator `%%`.

Solution 1: a naive `for` loop
```{r}
vec <- c()
for (n in 1:100){
  if (n %% 3 == 0){ vec[n] <- "Fizz" } 
  else if (n %% 5 == 0){ vec[n] <- "Buzz" } 
  else if (n %% 3 == 0 & n %% 5 == 0){ vec[n] <- "FizzBuzz" } 
  else { vec[n] <- n }
}

print(vec)
```


Solution 2: a better `for` loop

```{r eval=FALSE}
output <- vector()

for (i in 1:100) {
  output[i] <- ""
  if (i %% 3 == 0) {output[i] <- paste(output[i], "Fizz")}
  if (i %% 5 == 0) {output[i] <- paste(output[i], "Buzz")}
  if (output[i] == "") {output[i] <- i}
}

print(output)
```

(Reference: Rory Spanton. (2020). *How to Solve the FizzBuzz Problem in R*. https://towardsdatascience.com/how-to-solve-the-fizzbuzz-problem-in-r-c62e7e6c959a)    

***
*Notes:*

(1) Solution 1: Create an empty vector. Mark multiples of 3 as "Fizz". Mark multiples of 5 as "Buzz". Mark multiples of 3 and 5 as "FizzBuzz". Mark the remaining elements as the original numbers. The resulting vector is a character vector.

(2) Solution 2: Create an empty vector. Mark all elements as an empty string `""`. Inside the loop, `if (i %% 3 == 0) {output[i] <- paste(output[i], "Fizz")}` is evaluated first. If an element is a multiple of 3, "Fizz" is pasted to an empty string, resulting in "Fizz". Next, `if (i %% 5 == 0) {output[i] <- paste(output[i], "Buzz")}` is evaluated, which works in the same way as the first conditional statement. Lastly, if an element is a multiple of both 3 and 5, "Buzz" will be pasted to "Fizz". Remaining elements are marked by the original numbers and coerced to characters.


***
## Nested loops
A nested loop is a loop inside a loop. The "inner loop" will be executed one time for each iteration of the "outer loop".

***
*Example:* 

Write an R program which takes two digits m (row) and n (column) as input and generates a two-dimensional array. The element value in the i-th row and j-th column of the array should be i*j (i = 1.., m; j = 1.., n).

```{r}
m <- 3
n <- 4

row <- 1:m
col <- 1:n
a <- matrix(m*n, nrow = m, ncol = n)
for (i in row){
  for (j in col){
    a[i,j] <- i*j
  }
}

print(a)
```

*** 
*Notes:*

If `m=3` and `n=4`, we get a matrix of 3 rows and 4 columns. 

i      j      i*j     
-----  -----  -----  
1      1      1        
1      2      2        
1      3      3       
1      4      4     
2      1      2        
2      2      4        
2      3      6        
2      4      8        
3      1      3        
3      2      6        
3      3      9        
3      4      12      


***
## `while` loops
`while` loops repeat an expression while a condition is true.

***
*How it works:*
```
while (condition) expression
```

***
```{r}
i <- 5
while (i <= 25) {
  print(i)
  i <- i + 5
}
```

***
*Note:* We can rewrite any `for` loop with `while` instead. This means `while` is more flexible than `for`. So we should use `for` wherever possible. It’s good practice to use the least flexible solution to a problem.

***
*Example:* 

Write an R program to guess a number between 1 to 9. Hint: User is prompted to enter a guess. If the user guesses wrong then the prompt appears again until the guess is correct. On successful guess, the program will exit.

```{r eval=FALSE}
target_num <- sample(1:9, 1)
guess_num <- 0

while (guess_num != target_num){
  guess_num <- as.numeric(readline("Guess a number between 1 and 10 until you get it right : "))
}
```

***
## `break`
The `break` statement can be used to terminate a loop.

***
*Example:* 

Exit the loop if the number is equal to 15.
```{r}
i <- 5
while (i <= 25) {
  print(i)
  if (i == 15) break
  i <- i + 5
}
```



***
## `next`
The `next` statement can be used to discontinue one cycle and skip to the "next" cycle.

***
*Example:* 

Print all the numbers from 0 to 6 except 3 and 6.
```{r}
num <- 0:6
for (i in num) {
  if (i == 3 | i ==6) {
    next
  }
  print(i)
}
```
