---
title: "Reshaping Data between Long and Wide Formats"
author: "Yun Dai"
date: "04/2020"
output: 
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    toc_float: true
    css: contents.css
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
rm(list = ls())
library(readr)
long <- read_csv("/Volumes/GoogleDrive/My Drive/websites/new web/data/r/sample/long.csv")
```

******
In this post, we show two ways of converting datasets between long and wide formats: (1) using functions `spread()` and `gather()` from package `tidyr`; (2) using functions `dcast()` and `melt()` from package `data.table`.

******
## `tidyr`
```{r message=FALSE, warning=FALSE}
library(tidyr)
library(dplyr)
```

******
### long to wide
Dataset `long` is currently in the long format. We will reshape it to the wide format with `spread()`.
```{r}
long
```

******
`spread()` converts data from long to wide, spreading a key-value pair across multiple columns. The trick is to set the right `key` variable and the `value` variable. 

Below we add two new variables that are `gpa` of `course` A and B. Here, `key` is `course` and `value` is from `gpa`. 

```{r}
wide1 <- long %>% spread(key = course, value = gpa)
wide1
```

We can rename the two new columns if we want to.
```{r}
names(wide1)[4:5] <- c("gpa_A", "gpa_B")
wide1
```

******
Recently, `tidyr` has updated `spread()` to `pivot_wider()`, which provides more intuitive syntax and more options. `spread()` is not going away but will not be under active development.

```{r}
wide2 <- long %>% pivot_wider(names_from = semester, values_from = participation)
names(wide2)[4:5] <- c("participation_Fall2019", "participation_Spring2019")
wide2
```

******
If there are multiple variables whose values will be paired with the key variable, we can specify them with `c()` in the `values_from` argument.
```{r}
wide3 <- long %>% pivot_wider(names_from = course, values_from = c(gpa, participation))
wide3
```


******
### wide to long
`gather()` converts data from wide to long format for demo. 

We will use the dataset `wide1` in the wide format generated above.
```{r}
wide1 
```

Below we convert the dataset `wide1` back to the long format with `gather()`. Here the `key` variable to be created is `course`, and the `value` variable to be created is `gpa`. We remove the rows with `NA` in the value column.
```{r}
wide1 %>% gather(key = "course", value = "gpa", gpa_A, gpa_B, na.rm = T)
```

A bit more postprocessing on variable `course` to separate the course names `A` and `B` from the prefix `gpa_`.
```{r}
wide1 %>% 
  gather(key = "course", value = "gpa", gpa_A, gpa_B, na.rm = T) %>%
  separate(col = course, into = c("gpa_", "course"), sep = 4) %>%
  select(-gpa_)
```

******
Recently, `tidyr` has updated `gather()` to `pivot_longer()`. Like `pivot_wider()`, it provides more intuitive syntax and more options. `gather()` is not going away but will not be under active development.

```{r}
wide2
```

Using `pivot_longer()`, reshaping and postprocessing of column values can be done within itself. With arguments of `names_prefix`, `names_sep` or `names_pattern`, we no longer need `separate()` or `extract()`.
```{r}
wide2 %>% 
  pivot_longer(names_to = "semester", 
               values_to = "participation", 
               cols = participation_Fall2019 : participation_Spring2019,
               values_drop_na = T,
               names_prefix = "participation_")
```

******
## `data.table`
Package `data.table` also offers ways to convert the data between long and wide format.

```{r message=FALSE, warning=FALSE}
library(data.table)
```

******
### long to wide
`dcast()` reshapes datasets from long to wide formats. 

What we do below is the same as `long %>% spread(key = course, value = gpa)`.
```{r}
dcast(long, id + semester + participation ~ course, value.var = "gpa")
```

******
### wide to long
`melt()` reshapes data from wide to long. Instead of working on a key-value pair, it uses `id` and `measure`. 

The example below is the same as `wide1 %>% gather(key = course, value = gpa, -id, -semester, -participation)`.
```{r}
melt(wide1, id = c("id", "semester", "participation"), measure = c("gpa_A", "gpa_B"))
```

We can also remove the rows with `NA` by setting `na.rm = TRUE`.
```{r}
melt(wide1, id = c("id", "semester", "participation"), measure = c("gpa_A", "gpa_B"), na.rm = TRUE)
```

