---
title: "Writing R Code"
author: "Yun Dai"
date: "01/2019"
output: 
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    toc_float: true
    css: contents.css
---

******
## Being reproducible
Usually, when we work on a project, we will want to keep track of what we do. Below we discuss the best practices for writing R code to the goal of being reproducible, together with some general tips and tricks.

******
### setting up current working directory
The first thing we want to do is set up a working directory, which is the default directory to read and write files. 

`getwd()` returns the location of the current working directory. 

`setwd()` sets the current working directory. Once we set the working directory, we can access the files via relative path, which is much easier to specify than the absolute path.
```{r results=FALSE, message=FALSE, eval=FALSE}
setwd("~/Desktop")
```

******
### organizing files
For each project we will want to set up a different working directory, and put data, code, graphics, tables, documentations, reports and other files of one project in one folder.

******
## Organizing code
The recommended way to organize code is as below:

* Starting the script with the project information using comments
* Loading all required packages
* Sourcing codes
* Functions for specific tasks

```{r results=FALSE, message=FALSE, eval=FALSE}
######################################
# project description
######################################
# Author:
# Date:
# ...

######################################
# load libraries
######################################
library(readr)
library(dplyr)
library(stringr)
library(depmixS4)
library(ggplot2)

######################################
# run scripts
######################################
source("cleaning.R")
source("models.R")
source("viz.R")
######################################

```

******
### using comments
`#` indicates comments. Everything following `#` will be ignored.

We use comments not only to "comment", but also to

* keep track of our reasoning and let the readers (e.g. our collaborators) know what we have been doing
* break long scripts into blocks
* comment out commands to test parts of code in debugging

If we need to comment out a block of codes, we need to put `#` at the start of each new line.
```{r}
# this is 
# a very long
# comment
```

******
### intermediate objects
Defining intermediate objects as we move along gives the readers (and yourself when you later review your code) a clear clue of what you have done. It also makes debugging and checking intermediate outputs easier.

As an oversimplified example, below we first defined all the rules, logical expressions indicating elements or rows to keep, before passing them to `subset()`.
```{r results=FALSE, message=FALSE, eval=FALSE}
a <- rule1
b <- rule2
c <- rule3

subset(data, a & b & c)
```

******
### automating your code
If you find yourself repeating your code, automate them with a loop or a function. Doing so would also reduce chances of making mistakes in our code.

******
## Syntax
### spacing around binary operators (=, <-, + etc.) and after a comma
The exception is around `:` and `::`.

```{r results=FALSE, eval=FALSE}
x <- (-b + sqrt(b ^ 2 - 4 * a * c)) / (2 * a)
```

******
### assignment
Use `<-` for assignment instead of `=`.

******
### using consistent naming conventions
Whichever naming convention (underscore_separated, period.separated, CamelCase ...) we choose, it is best that we keep it consistent. 

******
## Miscellaneous

******
### conflicts in function names
Sometimes there could be conflicts in function names. For instance, packages `dplyr` and `plyr` have several functions that share the same names: `mutate()`, `rename()`, `summarise()` etc. We need to specify which package needs to be loaded when calling those functions (e.g.  `dplyr::summarise()` or `plyr::summarise()`).
 
******
## Read more
* [Google's R Style Guide](https://google.github.io/styleguide/Rguide.xml)
* [Style guide](http://stat405.had.co.nz/r-style.html)
* [Best Practices for Writing R Code](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/index.html)

